Modelling BCG vaccination in the UK

### 2019-08-05 rmds/09-model-fitting.Rmd
H model fitting and results comments

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets.

This chapter details an approach to challenging a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad-hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

State space models (SSMs) may be used to model dynamic systems that are partially observed (such as all but the most contained infectious disease outbreak or endemic). They consist of a set of parameters, a latent continuous/discrete time state process and an observed continuous/discrete time process.[@Murray2015] The model developed in the previous section represents the state process of the SSM, with the parameters estimated for the model representing the model initial conditions and parameter set. To complete the mapping to an SSM an observational model is required. This observational model takes the latent estimates from the dynamic model and forecasts the observed data. We specify such an observational model in the Section \@ref(observational-model).

@@

There are three major considerations to account for when developing an observed disease notification model (i.e a reporting model). These are: systematic reporting error over time; systematic changes in reporting error over time; and reporting noise. I assumed that all reporting errors are Gaussian and that their are no time variable reporting errors. This model was used for all data fitted to. The reporting model can be defined as follows,

@@

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000 (see Chapter \@ref(data)). A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, this may result in overfitting and so has not been implemented here.

@@

The model outlined in Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful information in the observed data may be lost.

@@

### Introduction

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates. Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

Libbi was used for all model fitting.[@Murray2015] LibBi is a software package for state-space modelling and Bayesian inference. It uses a domain specific language for model specification, which is then optimised and compiled to provide highly efficient model code. It focuses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across multiple CPUs or GPUs. `RBI` and `RBI.helpers` were used to interface with LibBi from R.[@Funk:2019ud; @Funk:2019uw] `RBI.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked, with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

1. For a given parameter set, the particle filter is initialised by drawing a number of random samples (state particles) from the initial conditions of the model under consideration. These samples are then given a uniform weighting.

@@

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models, a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate Bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC).[@Murray2015] This algorithm is used for all dynamic model fitting in this thesis.

@@

The accuracy of the marginal likelihood estimate returned by the particle filter is dependent on the number of particles used, the number of observed data points, the parameter sample, and the complexity of the model. As the number of particles tends towards infinity the likelihood estimate provided by the particle filter should tend towards the exact solution. This suggests that choosing a very high number of particles maybe the optimal solution in terms of accuracy. Unfortunately, each particle requires a full model simulation, which for complex models can be computationally costly. This means that using very large numbers of particles is not tractable. For this reason it is necessary to determine an optimal number of particles that both provides an adequately accurate estimate of the likelihood whilst being computationally tractable.

@@

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via LibBi,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 1024 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be in-feasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 1024 was specified as the maximum number of feasible particles to use in the particle filter.

@@

When using an MCMC algorithm a proposal distribution is required to provide new parameter samples to evaluate. For SMC-SMC a proposal distribution is required to inform the MCMC sampler that is run during the rejuvenation step. By default if no proposal distribution is provided Libbi uses the prior distribution.[@Murray2015] The prior distribution can be an inefficient proposal distribution as it is likely to have a low acceptance rate (from the MCMC sampler).[@Murray2015] Having a low acceptance rates means that many more MCMC steps are required to generate a successful parameter sample. This results in slow mixing and computationally expensive MCMC steps may make model fitting intractable.

@@

Understanding the impact of parameter variation can help when interpreting findings from a model, targeting interventions, and identifying parameters for which improved estimates are needed. Often parameter sensitivity is assessed using single-parameter or local sensitivity analyses. Unfortunately, these techniques do not accurately capture uncertainty or sensitivity in the system as they hold all other parameters fixed.[@Marino2009a] Various techniques exist that can globally study a multi-dimensional parameter space. In this section, I will outline the partial rank correlation coefficient method (PRCC) and discuss it's strengths and weaknesses. The implementation of PRCC to estimate the parameter sensitivity of parameters fitted using the model fitting pipeline presented above is then discussed.

@@

1. Model initialisation using minimal error checking and single precision computation. Implemented using the `disable-assert` and `enable-single` flags in LibBi.[@Murray2015] Outputs are only given for times with observed data and a subset of parameters are recorded for final reporting^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi].

2. 2000 parameter sets were taken from the prior distribution and the model was then simulated for each one.

3. Maximum likelihood optimisation with 100 steps, using the Nelder-Mead simplex method, via LibBi.[@Murray2015] This approximates the mean of the posterior distribution.

@@

7. SMC-SMC model fitting with 2000 initial parameter particles. Particle rejuvenation was set to trigger when the effective sample size decreased below 25%, with 10 MCMC steps used each time.

@@

After development of the model fitting pipeline but before results could be produced both the `optimise` (from `rbi`) and `adapt_particle` (from `rbi.helpers`) began to error with a difficult to debug `LibBi` error message. This meant that steps 3-5 of the model fitting pipeline could not be used with the final model. As a work around the maximum permitted number of particles (1024) was used for all model fitting (increasing the number of particles was also explored as detailed in the previous section).

As discussed, there were multiple issues with fitting the final model and this made it diff ult to determine what the mean of posterior distribution was. This made manually tuning the proposal distribution challenging and so instead a standard deviation of 1% was used. This value was chosen as it increased the acceptance rate by limiting each rejuvenation step to a relatively small subset of the prior distribution whilst not preventing the exploration of new parameter space in scenarios when model fits from the initial particle sample were poor.

@@

### Model Fit to TB incidence rates from the ETS

The fitted model consistently under-predicted overall TB incidence for all years with data (Table \@ref(tab:09-table-incidence-preds)). It also failed to capture the overall trend in TB incidence rates with the forecast incidence increasing year-on-year in comparison to the observed data which showed greater variation. Stratifying by age shows that the model also failed to captured the age distribution of TB incidence (Figure \@ref(fig:09-age-strat-incidence). Whilst the model under predicted TB incidence in all age-groups it was more accurate for older adults, implying that even if the magnitude of cases had been better predicted the distribution would still not match the observed data.

```{r 09-table-incidence-preds, message = FALSE, warning = FALSE}
sum_obs_table %>% 
  set_names(., nm = c(names(.)[-ncol(.)], "Predicted Incidence (95% CI)")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Observed versus predicted overall TB incidence for years that the model was fitted too.", 
  caption = "Observed versus predicted overall TB incidence for years that the model was fitted too. The model under-predicted incidence for all years fitted too. The upper credible interval of the model predictions was substantially lower than the observed data. (95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

@@

Figure \@ref(fig:09-fig-sens) shows the partial rank correlation coefficients for each parameter that was fitted too. This indicates that variation in the historic effective contact rate leads to the greatest variation in TB incidence. Based on the model structure this makes sense as small changes in the historic contact rate will compounded over the run time of the model. Similarly the half-life of the historic contact rate also plays a dominant role. The non-UK born mixing modifier for young adults leads to the second largest change in TB incidence. This corresponds to the findings from the scenario analysis discussed above in which the introduction if this parameter resulted in a greatly improved model fit. Surprisingly, the transmission probability modifier has relatively little impact on TB incidence compared to the other parameters fitted to, disagreeing with the findings from the scenario analysis. The lack of diversity in the posterior distribution seen above means that these findings are indicative only.

@@

As discussed earlier in this chapter, SMC-SMC was used for all model fitting. This uses an external SMC step to estimated the posterior distribution as well as an internal SMC step to estimate the marginal likelihood. In general Bayesian model fitting approaches are beneficial as they allow prior information to be fully incorporated into model fitting and they produce posterior distribution estimates rather than point estimates.[@Murray2015; @Toni2009] There are two main families of approaches, SMC and MCMC, both previously discussed in this chapter. Using SMC to estimate the posterior distribution has numerous advantages over MCMC. The first of these is that MCMC approaches are sensitive to their initial conditions. If a model has multiple local best fits MCMC may only converge to a single minima rather than fully exploring the posterior distribution. Multiple MCMC chains may be used to try and account for this but as each chain must be independently run to convergence only a few concurrent chains are likely to be practical. SMC on the other hand is initialised with a large sample from the prior distribution, meaning that local minimas are more likely to be explored. Parameter particle weighting and re-sampling then balances the contribution to the posterior distribution of these local minimas based on their fit to the observed data. Additionally, MCMC approaches are by definition sequential,[@Murray2015] although if they make use of particle filters these can be run in parallel. Increasing the number of particles in a filter may lead to an increase in the chain mixing rate of the MCMC chain but as particles numbers are increased any returns will decrease. To account for this multiple chains are often used, but as outlined above the burn-in required for each chain limits the potential speed-up. In comparison, each SMC parameter particle can have it's marginal likelihood computed separately. Although the re-sampling step remains a bottleneck as it can only be completed once all marginal likelihoods have been computed. On the other hand SMC is less interactive than MCMC meaning that model fitting is harder to inspect when it is in progress. This is because SMC is not sequential, unlike an MCMC run for which each draw can be inspected as it is computed. Similarly, as SMC is not a sequential technique multiple runs cannot be combined. This means that model fitting must be done in a single run using a priori knowledge to judge the number of MCMC rejuvenation steps required, and the expected total run time. SMC will also have a variable run time based on the effective sample size as rejuvenation only happens when parameter particles have been depleted beyond a certain point. An additional benefit of SMC approaches are that they can theoretically be extended to model selection as well as parameter posterior estimation - effectively estimating posteriors for candidate model structures.[@Toni2009] Beyond SMC and MCMC there are multiple other model fitting algorithms that each have their own strengths and benefits - discussion of these is outside the scope of this thesis.

@@

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Indeed the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's `R` interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be extremely diff ult to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development.

@@

* Defined the disease transmission model from the previous chapter as a state-space model. 
* Outlined the observed data and defined an appropriate measurement model for this data. 
* Developed a model fitting pipeline based on SMC-SMC to fit the previously defined state-space model to TB notification data.
* Discussed ad-hoc approaches used to improve model fit. 
* Evaluated the various scenarios discussed in the previous chapter using DIC and discussed the implications of these findings. 
* Presented the model fit to data from the best fitting scenario as established using the DIC. 
* Discussed the posterior distributions of the fitted parameters and compared these distributions to the prior distributions.
* Estimated the model sensitivity to parameter changes and discussed the impact that this sensitivity may have.
* Discussed the strengths and limitations of the work presented here, as well as outlining potential further work.

>>>>

In the previous chapter I outlined a mechanistic model of Tuberculosis (TB) transmission. Whilst this model made use of the best available evidence there remains a large degree of uncertainty regarding its structure and parameterisation. The majority of this uncertainty relates to the amount of TB transmission occurring in England. In order to use this model to understand TB transmission, and the impact of different BCG vaccination policies (Chapter \@ref(model-results)) this uncertainty needs to be reduced and the parameter space tightened to reflect more realistic ranges. An approach to deal with this uncertainty is to fit the model to available observed data. Model fitting involves optimising over the available parameter space to return parameter sets that fit the data in some quantitative way "better" than other parameter sets. An alternative to model fitting is using a model parameterised with expert knowledge only for inference. This approach is not appropriate here due to the large amount of uncertainty for many of the model parameters. Any inference based in just the priors would have extremely large credible intervals, reflect reality poorly, and likely be biased in multiple areas.

This chapter details an approach to fitting a infectious disease model to data using the state-space model formulation and Bayesian model fitting techniques. It first outlines the infectious disease model discussed in the previous chapter as a state-space model, as well as detailing the data used for fitting the model, and the parameters that are fitted. It then outlines the theoretical, and practical, justification for the model fitting pipeline used to calibrate and fit this state space model. Finally it discusses the quality of the model fit, ad-hoc techniques used to improve model fit, strengths and limitations of the approach, and areas for further work.

@@

State space models (SSMs) may be used to model dynamic systems that are partially observed (such as all but the most contained infectious disease outbreak or endemic). They consist of a set of parameters, a latent continuous/discrete time state process and an observed continuous/discrete time process.[@Murray2015] The model developed in the previous section represents the state process of the SSM, with the parameters estimated for the model representing the model initial conditions and parameter set. To complete the mapping to an SSM an observational model is required. This observational model takes the latent estimates from the dynamic model and forecasts the observed data. Ispecify such an observational model in the Section \@ref(observational-model).

@@

There are three major considerations to account for when developing an observed disease notification model (i.e a reporting model). These are: systematic reporting error over time; systematic changes in reporting error over time; and reporting noise. I assumed that all reporting errors are Gaussian and that there are no time variable reporting errors. This model was used for all data fitted to. The reporting model can be defined as follows,

@@

A potential limitation of this model is that reporting of TB cases is likely to have improved over time. This is especially true of notifications reported prior to the introduction of the ETS in 2000. A potential improvement to this model would be to introduce separate systematic reporting errors for notifications pre and post the introduction of the ETS. However, this may result in overfitting and so has not been implemented here.

@@

The model outlined in Chapter \@ref(model-development) has a large number of free parameters for which prior distributions have been specified based on the observed data, the literature, and expert knowledge. In theory the model fitting pipeline outlined below could be used to produce posterior distributions for all these parameters. However, in practice this is not feasible as the data discussed in Section \@ref(observed-data) only covers notifications and therefore does not contain sufficient information. If every parameter was allowed to update based on the data then it is likely that the resulting posterior distributions would not match with alternative data sources and the literature. Another potential issue is that by allowing all parameters to be fitted the meaningful transmission related information in the observed data may be lost due to overfitting from other variables.

@@

Fitting dynamic transmission models is complex and requires the use of specialist statistical techniques. There are a variety of these tools available. Ranging from tried and tested to cutting edge. Historically many modellers have used maximum likelihood methods to fit deterministic models. Unfortunately, whilst computationally efficient, these methods only provide point estimates and are limited to relatively simple models. More recently Bayesian methods have become popular. These have numerous benefits including: explicit inclusion of prior knowledge via prior distributions for all parameters; ability to handle complex stochastic models; and provide parameter distributions (posterior distribution) of best fitting parameters rather than single point estimates.[@Murray2015] Unfortunately these methods also require calibration prior to use. This section outlines the theoretical justification, and implementation details, of an automated model fitting pipeline used to fit the previously detailed state space model.

`LibBi` was used for all model fitting.[@Murray2015] `LibBi` is a software package for state-space modelling and Bayesian inference. It uses a domain specific language for model specification, which is then optimised and compiled to provide highly efficient model code. It focuses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across multiple CPUs or GPUs. The `rbi` and `rbi.helpers` packages were used to interface with `LibBi` from R.[@Funk:2019ud; @Funk:2019uw] `rbi.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked (using CPU voltage manipulation), with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. Whilst a compute cluster was theoretically available, in practise the hardware available was limited, installing `LibBi` was challenging, and run times were constrained by fair access. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

1. *Sampling*: For a given parameter set, the particle filter is initialised by drawing a number of random samples (state particles) from the initial conditions of the model under consideration. These samples are then given a uniform weighting.

@@

 * *Re-sampling:* The particle stocke is restorted to equal weights by re-sampling particles, with replacement, with the probability of each sample being drawn being proportional to its weight.

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models, a Kalman filter may be used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate Bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC).[@Murray2015] This algorithm is used for all dynamic model fitting in this thesis.

@@

The accuracy of the marginal likelihood estimate returned by the particle filter is dependent on the number of particles used, the number of observed data points, the parameter sample, and the complexity of the model. As the number of particles tends towards infinity the likelihood estimate provided by the particle filter should tend towards the exact solution. This suggests that choosing a very high number of particles may be the optimal solution in terms of accuracy. Unfortunately, each particle requires a full model simulation, which for complex models can be computationally costly. This means that using very large numbers of particles is not tractable. For this reason it is necessary to determine an optimal number of particles that both provides an adequately accurate estimate of the likelihood whilst being computationally tractable.

@@

I have implemented this as a two step process for each fitted scenario. Firstly, I used the Nelder-Mead simplex method, via `LibBi`,[@Murray2015] to find a parameter set that optimised the maximum likelihood. I then initialised a 1000 step PMCMC chain with this parameter set, using 1024 particles in the particle filter. I then used `rbi.helpers`,[@Funk:2019uw] as outlined above, to estimate the number of particles required to produce a log-likelihood variance of less than 1 for this sample of the posterior distribution, using 250 samples per step and starting with 16 particles. I initially planned to repeat this process for multiple draws from the posterior distribution but this proved to be in-feasible given the compute available. A target of 5 for the log-likelihood variance was chosen as a smaller target could not be feasibly achieved given the compute resources available. Additionally 1024 was specified as the maximum number of feasible particles to use in the particle filter.

@@

When using an MCMC algorithm a proposal distribution is required to provide new parameter samples to evaluate. For SMC-SMC a proposal distribution is required to inform the MCMC sampler that is run during the rejuvenation step. By default if no proposal distribution is provided `LibBi` uses the prior distribution.[@Murray2015] The prior distribution can be an inefficient proposal distribution as it is likely to have a low acceptance rate (from the MCMC sampler).[@Murray2015] Having a low acceptance rates means that many more MCMC steps are required to generate a successful parameter sample. This results in slow mixing and computationally expensive MCMC steps may make model fitting intractable.

@@

Understanding the impact of parameter variation can help when interpreting findings from a model, targeting interventions, and identifying parameters for which improved estimates are needed. Often parameter sensitivity is assessed using single-parameter or local sensitivity analyses. Unfortunately, these techniques do not accurately capture uncertainty or sensitivity in the system as they hold all other parameters fixed.[@Marino2009a] Multiple techniques exist that can globally study a multi-dimensional parameter space but the partial rank correlation coefficient method (PRCC) that I will discuss, and implement, here has been shown to be both reliable and efficient.[@Marino2009a]

@@

1. Model initialisation using minimal error checking and single precision computation. Implemented using the `disable-assert` and `enable-single` flags in `LibBi`.[@Murray2015] Outputs are only given for times with observed data and a subset of parameters are recorded for final reporting^[Model code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/inst/bi/BaseLineModel.bi].

2. 1000 parameter sets were taken from the prior distribution and the model was then simulated for each one.

3. Maximum likelihood optimisation with 100 steps, using the Nelder-Mead simplex method, via `LibBi`.[@Murray2015] This approximates the mean of the posterior distribution.

@@

7. SMC-SMC model fitting with 1000 initial parameter particles. Particle rejuvenation was set to trigger when the effective sample size decreased below 25%, with 10 MCMC steps used each time.

@@

After development of the model fitting pipeline but before results could be produced both the `optimise` (from `rbi`) and `adapt_particle` (from `rbi.helpers`) began to error with multiple, transient, `LibBi` error messages. This meant that steps 3-5 of the model fitting pipeline could not be used with the final model. As a work around the maximum permitted number of particles (1024) was used for all model fitting (increasing the number of particles was also explored as detailed in the previous section).

As discussed, there were multiple issues with fitting the final model and this made it difficult to determine what the mean of posterior distribution was. This made manually tuning the proposal distribution challenging and so instead a standard deviation of 1% was used. This value was chosen as it increased the acceptance rate by limiting each rejuvenation step to a relatively small subset of the prior distribution whilst not preventing the exploration of new parameter space in scenarios when model fits from the initial particle sample were poor.

@@

### Model Fit to TB incidence from the ETS

The fitted model consistently under-predicted overall TB cases for all years with data (Table \@ref(tab:09-table-incidence-preds)). It also failed to capture the overall trend in TB incidence with the forecast incidence increasing year-on-year in comparison to the observed data which showed greater variation. Stratifying by age shows that the model also failed to captured the age distribution of TB incidence (Figure \@ref(fig:09-age-strat-incidence). Whilst the model under predicted TB incidence in all age-groups it was more accurate for older adults, implying that even if the magnitude of cases had been better predicted the distribution would still not match the observed data.

```{r 09-table-incidence-preds, message = FALSE, warning = FALSE}
sum_obs_table %>% 
  set_names(., nm = c(names(.)[-ncol(.)], "Predicted Incidence (95% CI)")) %>% 
   set_names(., nm = names(.) %>% str_replace_all("Incidence", "Cases")) %>% 
  select(-Observation) %>% 
kable(
  caption.short = "Observed versus predicted overall TB cases for years that the model was fitted to.", 
  caption = "Observed versus predicted overall TB cases for years that the model was fitted to. (95\\% CI): 95\\% credible interval estimated using the 2.5\\% quantile and the 97.5\\% quantile.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(latex_options = c("hold_position")) %>% 
  {.}
```

@@

Figure \@ref(fig:09-fig-sens) shows the partial rank correlation coefficients for each parameter that was fitted too. This indicates that variation in the historic effective contact rate leads to the greatest variation in TB incidence. Based on the model structure this makes sense as small changes in the historic contact rate will compounded over the run time of the model. Similarly the half-life of the historic contact rate also plays a dominant role. The non-UK born mixing modifier for young adults leads to the second largest change in TB incidence. This corresponds to the findings from the scenario analysis discussed above in which the introduction of this parameter resulted in a greatly improved model fit. Surprisingly, the transmission probability modifier has relatively little impact on TB incidence compared to the other parameters fitted to, disagreeing with the findings from the scenario analysis. The lack of diversity in the posterior distribution seen above means that these findings are indicative only.

@@

As discussed earlier in this chapter, SMC-SMC was used for all model fitting. This uses an external SMC step to estimate the posterior distribution as well as an internal SMC step to estimate the marginal likelihood. In general Bayesian model fitting approaches are beneficial as they allow prior information to be fully incorporated into model fitting and they produce posterior distribution estimates rather than point estimates.[@Murray2015; @Toni2009] There are two main families of approaches, SMC and MCMC. Using SMC to estimate the posterior distribution has numerous advantages over MCMC. The first of these is that MCMC approaches are sensitive to their initial conditions. If a model has multiple local best fits MCMC may only converge to a single minima rather than fully exploring the posterior distribution. Multiple MCMC chains may be used to try and account for this but as each chain must be independently run to convergence only a few concurrent chains are likely to be practical. SMC on the other hand is initialised with a large sample from the prior distribution, meaning that local minimas are more likely to be explored. Parameter particle weighting and re-sampling then balances the contribution to the posterior distribution of these local minimas based on their fit to the observed data. Additionally, MCMC approaches are by definition sequential,[@Murray2015] although if they make use of particle filters these can be run in parallel. Increasing the number of particles in a filter may lead to an increase in the chain mixing rate of the MCMC chain but as particles numbers are increased any returns will decrease. To account for this multiple chains are often used, but as outlined above the burn-in required for each chain limits the potential speed-up. In comparison, each SMC parameter particle can have it's marginal likelihood computed separately. Although the re-sampling step remains a bottleneck as it can only be completed once all marginal likelihoods have been computed. On the other hand SMC is less interactive than MCMC meaning that model fitting is harder to inspect when it is in progress. This is because SMC is not sequential, unlike an MCMC run for which each draw can be inspected as it is computed. Similarly, as SMC is not a sequential technique multiple runs cannot be combined. This means that model fitting must be done in a single run using a priori knowledge to judge the number of MCMC rejuvenation steps required, and the expected total run time. SMC will also have a variable run time based on the effective sample size as rejuvenation only happens when parameter particles have been depleted beyond a certain point. An additional benefit of SMC approaches are that they can theoretically be extended to model selection as well as parameter posterior estimation - effectively estimating posteriors for candidate model structures.[@Toni2009] Beyond SMC and MCMC there are multiple other model fitting algorithms that each have their own strengths and benefits - discussion of these is outside the scope of this thesis.

@@

This chapter showcased the use of `LibBi`, `rbi`, and `rbi.helpers` for fitting a complex transmission model. Unfortunately, the quality of the model fit was poor and the complexity of `LibBi` makes understanding the root cause difficult. It is possible that the model developed in the last chapter is to complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time) than available.[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] Indeed the model itself may not be identifiable with the available data.[@Lintusaari2016] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's `R` interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be frustrating to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it difficult for newcomers to contribute towards its development. It is possible that the fitting issues outlined here may be due to errors in the code used to implement the model from the previous chapter. This is unlikely as manual prior tuning has shown that a relatively good fit to the observed data can be achieved with the current model but these results could not be reproduced using the model fitting pipeline. However, this does not mean that the modelling code is bug free. Model bugs, if present, would bias results from the fitted model rather than preventing a good fit to the data or would prevent the model from being manually tuned to fit the data.

@@

* Defined the disease transmission model from the previous chapter as a state-space model, outlined the available data to be used for fitting it, and detailed a measurement model to link the observed data with the dynamic TB model.
* Developed a model fitting pipeline based on SMC-SMC to fit the previously defined state-space model to TB notification data. The theoretical background for this approach was outlined as well as the key steps for implementing it in practise. 
* As the quality of model fit achieved was poor ad-hoc calibration approaches, and the results they gave, were discussed.
* The scenarios outlined in the previous chapter were evaluated using DIC and the implications of the findings were discussed. 
* Model forecasts from the best fitting scenario, as established using the DIC, were compared to observed data. Posterior distributions from this model fit were then contrasted with the prior distributions. Finally, parameter sensitivity was estimated using the posterior distributions.
* Discussed the strengths and limitations of the work presented here, as well as outlining potential further work.
